---
title: "01_gradient_data_prep"
output: html_document
date: "2023-05-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Goal: Figure out which samples and EMPO3 categories are going to be used for the gradient analysis
Want to have apples to apples, by taking the lowest common sample number within empo category across plots

For example, if all plots have at least 3 instances of plant corpus, even if one plot has 7, just take 3 (at random).

```{r}
library(dplyr)
```

```{r}

# import data
fung_abun <- readRDS("../../intermediates/global/fully_filtered_p20_fungal_otu_table_matched_up.rds")
bact_abun <- readRDS("../../intermediates/global/fully_filtered_p20_bact_otu_table_matched_up.rds")

fung_meta <- readRDS("../../intermediates/global/fully_filtered_p20_fungal_otu_metadata_matched_up.rds")
bact_meta <- readRDS("../../intermediates/global/fully_filtered_p20_bact_otu_metadata_matched_up.rds")

# some data formatting
fung_meta$trophic[!(fung_meta$trophic %in% c("PrimaryProducer", "Environmental"))] <- "Consumer"
bact_meta$trophic[!(bact_meta$trophic %in% c("PrimaryProducer", "Environmental"))] <- "Consumer"

fung_dummy <- fung_abun[row.names(fung_abun)=="dummy",]
bact_dummy <- bact_abun[row.names(bact_abun)=="dummy",]

names(fung_dummy) <- fung_meta$sample_id[match(names(fung_abun), fung_meta$x_seq_id)]
names(bact_dummy) <- bact_meta$sample_id[match(names(bact_abun), bact_meta$x_seq_id)]


fung_abun <- fung_abun[row.names(fung_abun)!="dummy",]
bact_abun <- bact_abun[row.names(bact_abun)!="dummy",]



```

# Fungi 

```{r}

fung_meta_terr <- fung_meta[fung_meta$habitat=="Terrestrial",]
saveRDS(fung_meta_terr, "../../intermediates/gradient/terrestrial/fung_terrestrial_metadata.rds")

fung_abun_terr <- fung_abun[,names(fung_abun) %in% fung_meta_terr$x_seq_id]
fung_abun_terr <- fung_abun_terr[rowSums(fung_abun_terr)>0,] # 1911 OTUs, 718 samples

names(fung_abun_terr) <- fung_meta_terr$sample_id[match(names(fung_abun_terr), fung_meta_terr$x_seq_id)]


fung_terr_samp_counts <- fung_meta_terr %>% dplyr::group_by(empo_3, site_name) %>% dplyr::summarise(sum = n())

fung_terr_min_df <- fung_terr_samp_counts %>% dplyr::group_by(empo_3) %>% dplyr::slice(which.min(sum))


```


Okay I have the minimum sample counts, now to take x samples from each empo3 category in a plot...

```{r}


empos <- unique(fung_terr_min_df$empo_3)
# need to remove animal distal gut and animal surface because these are not present in Ridge site
empos <- empos[!empos %in% c("Animal distal gut", "Animal surface")]

# update min count df
fung_terr_min_df <- fung_terr_min_df[fung_terr_min_df$empo_3 %in% empos,]


plots <- unique(fung_meta_terr$site_name)

take_random <- function(empo, abun, meta, min_count_df) {
  sub_met <- meta[meta$empo_3==empo,]
  
  sub_abun <- abun[,names(abun) %in% sub_met$sample_id]
  sub_abun <- sub_abun[rowSums(sub_abun)>0,] 
  
  plots <- unique(sub_met$site_name)
  
  res <- list()
  
  for (a_plot in plots) {
    plot_met <- sub_met[sub_met$site_name==a_plot,]
    plot_abun <- sub_abun[,names(sub_abun) %in% plot_met$sample_id]
    plot_abun <- plot_abun[rowSums(plot_abun)>0,]
    
    min_empo_num <- min_count_df$sum[min_count_df$empo_3==empo]
    
    plot_abun <- plot_abun[,order(colSums(plot_abun), decreasing = TRUE)]
    
    random_samps_abun <- plot_abun[,sample(names(plot_abun), min_empo_num)]
    random_samps_abun <- random_samps_abun[rowSums(random_samps_abun)>0,]
    
    res[[a_plot]] <- random_samps_abun
    
  }
  
  return(res)
  
}


fung_empo_list <- lapply(empos,
                   take_random,
                   abun = fung_abun_terr,
                   meta = fung_meta_terr,
                   min_count_df = fung_terr_min_df)
names(fung_empo_list) <- empos


multi_merge_matrix <- function(matrix_list) {
  
  r <- merge(matrix_list[[1]], matrix_list[[2]], by="row.names", all=TRUE)
  # ASVs become a row, change back to rownames
  r <- column_to_rownames(r, "Row.names")

  for (i in 3:length(matrix_list)) {
  r <- merge(r, matrix_list[[i]], by="row.names", all=T)
  rownames(r) <- r[,1]
  r[,1] <- NULL
  }

  r[is.na(r)] <- 0

  return(r)

}

# animal distal gut and animal surface no more ridge


```

I have a list of otu tables for each empo category
Each empo category has a list of otu tables for each plot
I want one otu table per plot with all the empo categories to make networks

e.g. I want all the beach otu tables together

```{r}
 
### ENTRANCE
fung_entrance <- list()
for (i in 1:6) {
  fung_entrance[[i]] <- fung_empo_list[[i]][1]
  
}

fung_entrance_abun <- multi_merge_matrix(fung_entrance)


### CONFLUENCE
fung_confluence <- list()
for (i in 1:6) {
  fung_confluence[[i]] <- fung_empo_list[[i]][2]
  
}

fung_confluence_abun <- multi_merge_matrix(fung_confluence)

### WATERFALL
fung_waterfall <- list()
for (i in 1:6) {
  fung_waterfall[[i]] <- fung_empo_list[[i]][3]
  
}

fung_waterfall_abun <- multi_merge_matrix(fung_waterfall)

### DRUMROAD
fung_drumroad <- list()
for (i in 1:6) {
  fung_drumroad[[i]] <- fung_empo_list[[i]][4]
  
}

fung_drumroad_abun <- multi_merge_matrix(fung_drumroad)

### ESTUARY
fung_estuary <- list()
for (i in 1:6) {
  fung_estuary[[i]] <- fung_empo_list[[i]][5]
  
}

fung_estuary_abun <- multi_merge_matrix(fung_estuary)


### BEACH
fung_beach <- list()
for (i in 1:6) {
  fung_beach[[i]] <- fung_empo_list[[i]][6]
  
}

fung_beach_abun <- multi_merge_matrix(fung_beach)


### RIDGE
fung_ridge <- list()
for (i in 1:6) {
  fung_ridge[[i]] <- fung_empo_list[[i]][7]
  
}

fung_ridge_abun <- multi_merge_matrix(fung_ridge)


all_fung_plot_abun <- list(fung_entrance_abun,
                           fung_confluence_abun,
                           fung_waterfall_abun,
                           fung_drumroad_abun,
                           fung_estuary_abun,
                           fung_beach_abun,
                           fung_ridge_abun)


library(stringr)

# remove the plot name from samples
clean_sample_names <- function(df) {
  names(df) <- str_extract(names(df), '\\b\\w+$')
  return(df)
}

all_fung_plot_abun <- lapply(all_fung_plot_abun, clean_sample_names)

names(all_fung_plot_abun) <- plots

saveRDS(all_fung_plot_abun, "../../intermediates/gradient/terrestrial/six_empos_fung_filt_by_plot_list_for_networks.rds")





# add dummy row from prev20 culling, but only for samples existing in the empo/plot tables
all_fung_plot_abun <- readRDS("../../intermediates/gradient/terrestrial/six_empos_fung_filt_by_plot_list_for_networks.rds")

fung_samp_names <- lapply(all_fung_plot_abun, names)

fung_dummy_by_table <- lapply(fung_samp_names, 
                              function(x) as.data.frame(fung_dummy[,names(fung_dummy) %in% x]))


all_fung_with_dummy <- list()

for(i in 1:7) {
  attach <- rbind(all_fung_plot_abun[[i]],
                  fung_dummy_by_table[[i]])
  all_fung_with_dummy[[i]] <- attach
}


```



# Bacteria

```{r}

bact_meta_terr <- bact_meta[bact_meta$habitat=="Terrestrial",]
saveRDS(bact_meta_terr, "../../intermediates/gradient/terrestrial/bact_terrestrial_metadata.rds")


bact_abun_terr <- bact_abun[,names(bact_abun) %in% bact_meta_terr$x_seq_id]
bact_abun_terr <- bact_abun_terr[rowSums(bact_abun_terr)>0,] # 9615 OTUs, 718 samples

names(bact_abun_terr) <- bact_meta_terr$sample_id[match(names(bact_abun_terr), bact_meta_terr$x_seq_id)]


bact_terr_samp_counts <- bact_meta_terr %>% dplyr::group_by(empo_3, site_name) %>% dplyr::summarise(sum = n())

bact_terr_min_df <- bact_terr_samp_counts %>% dplyr::group_by(empo_3) %>% dplyr::slice(which.min(sum))
# update min count df
bact_terr_min_df <- bact_terr_min_df[bact_terr_min_df$empo_3 %in% empos,]

```



```{r}
# this is different from fungi because we want to use the same randomly selected samples as picked for the fungi
# so this function just gets otu tables by plot, which will be subset by sample id later
get_bact_plot_abun <- function(empo, abun, meta) {
  sub_met <- meta[meta$empo_3==empo,]
  
  sub_abun <- abun[,names(abun) %in% sub_met$sample_id]
  sub_abun <- sub_abun[rowSums(sub_abun)>0,] 
  
  plots <- unique(sub_met$site_name)
  
  res <- list()
  
  for (a_plot in plots) {
    plot_met <- sub_met[sub_met$site_name==a_plot,]
    plot_abun <- sub_abun[,names(sub_abun) %in% plot_met$sample_id]
    plot_abun <- plot_abun[rowSums(plot_abun)>0,]
    
    res[[a_plot]] <- plot_abun
    
  }
  
  return(res)
  
}


bact_empo_list <- lapply(empos,
                   get_bact_plot_abun,
                   abun = bact_abun_terr,
                   meta = bact_meta_terr)
names(bact_empo_list) <- empos


multi_merge_matrix <- function(matrix_list) {
  
  r <- merge(matrix_list[[1]], matrix_list[[2]], by="row.names", all=TRUE)
  # ASVs become a row, change back to rownames
  r <- column_to_rownames(r, "Row.names")

  for (i in 3:length(matrix_list)) {
  r <- merge(r, matrix_list[[i]], by="row.names", all=T)
  rownames(r) <- r[,1]
  r[,1] <- NULL
  }

  r[is.na(r)] <- 0

  return(r)

}

# animal distal gut and animal surface no more ridge


```

I have a list of otu tables for each empo category
Each empo category has a list of otu tables for each plot
I want one otu table per plot with all the empo categories to make networks

e.g. I want all the beach otu tables together

```{r}
 
### ENTRANCE
bact_entrance <- list()
for (i in 1:6) {
  bact_entrance[[i]] <- bact_empo_list[[i]][1]
  
}

bact_entrance_abun <- multi_merge_matrix(bact_entrance)
bact_entrance_abun <- bact_entrance_abun[,names(bact_entrance_abun) %in% names(fung_entrance_abun)]
bact_entrance_abun <- bact_entrance_abun[rowSums(bact_entrance_abun)>0,]


### CONFLUENCE
bact_confluence <- list()
for (i in 1:6) {
  bact_confluence[[i]] <- bact_empo_list[[i]][2]
  
}

bact_confluence_abun <- multi_merge_matrix(bact_confluence)
bact_confluence_abun <- bact_confluence_abun[,names(bact_confluence_abun) %in% names(fung_confluence_abun)]
bact_confluence_abun <- bact_confluence_abun[rowSums(bact_confluence_abun)>0,]

### WATERFALL
bact_waterfall <- list()
for (i in 1:6) {
  bact_waterfall[[i]] <- bact_empo_list[[i]][3]
  
}

bact_waterfall_abun <- multi_merge_matrix(bact_waterfall)
bact_waterfall_abun <- bact_waterfall_abun[,names(bact_waterfall_abun) %in% names(fung_waterfall_abun)]
bact_waterfall_abun <- bact_waterfall_abun[rowSums(bact_waterfall_abun)>0,]


### DRUMROAD
bact_drumroad <- list()
for (i in 1:6) {
  bact_drumroad[[i]] <- bact_empo_list[[i]][4]
  
}

bact_drumroad_abun <- multi_merge_matrix(bact_drumroad)
bact_drumroad_abun <- bact_drumroad_abun[,names(bact_drumroad_abun) %in% names(fung_drumroad_abun)]
bact_drumroad_abun <- bact_drumroad_abun[rowSums(bact_drumroad_abun)>0,]

### ESTUARY
bact_estuary <- list()
for (i in 1:6) {
  bact_estuary[[i]] <- bact_empo_list[[i]][5]
  
}

bact_estuary_abun <- multi_merge_matrix(bact_estuary)
bact_estuary_abun <- bact_estuary_abun[,names(bact_estuary_abun) %in% names(fung_estuary_abun)]
bact_estuary_abun <- bact_estuary_abun[rowSums(bact_estuary_abun)>0,]


### BEACH
bact_beach <- list()
for (i in 1:6) {
  bact_beach[[i]] <- bact_empo_list[[i]][6]
  
}

bact_beach_abun <- multi_merge_matrix(bact_beach)
bact_beach_abun <- bact_beach_abun[,names(bact_beach_abun) %in% names(fung_beach_abun)]
bact_beach_abun <- bact_beach_abun[rowSums(bact_beach_abun)>0,]


### RIDGE
bact_ridge <- list()
for (i in 1:6) {
  bact_ridge[[i]] <- bact_empo_list[[i]][7]
  
}

bact_ridge_abun <- multi_merge_matrix(bact_ridge)
bact_ridge_abun <- bact_ridge_abun[,names(bact_ridge_abun) %in% names(fung_ridge_abun)]
bact_ridge_abun <- bact_ridge_abun[rowSums(bact_ridge_abun)>0,]


all_bact_plot_abun <- list(bact_entrance_abun,
                           bact_confluence_abun,
                           bact_waterfall_abun,
                           bact_drumroad_abun,
                           bact_estuary_abun,
                           bact_beach_abun,
                           bact_ridge_abun)


library(stringr)

# remove the plot name from samples
clean_sample_names <- function(df) {
  names(df) <- str_extract(names(df), '\\b\\w+$')
  return(df)
}

all_bact_plot_abun <- lapply(all_bact_plot_abun, clean_sample_names)

names(all_bact_plot_abun) <- plots

saveRDS(all_bact_plot_abun, "../../intermediates/gradient/terrestrial/six_empos_bact_filt_by_plot_list_for_networks.rds")



# add dummy row from prev20 culling, but only for samples existing in the empo/plot tables
all_bact_plot_abun <- readRDS("../../intermediates/gradient/terrestrial/six_empos_bact_filt_by_plot_list_for_networks.rds")

bact_samp_names <- lapply(all_bact_plot_abun, names)

bact_dummy_by_table <- lapply(bact_samp_names, 
                              function(x) as.data.frame(bact_dummy[,names(bact_dummy) %in% x]))


all_bact_with_dummy <- list()

for(i in 1:7) {
  attach <- rbind(all_bact_plot_abun[[i]],
                  bact_dummy_by_table[[i]])
  all_bact_with_dummy[[i]] <- attach
}

```

```{r}
ff <- lapply(all_fung_with_dummy, function(x) x[,sort(names(x))])
saveRDS(ff, "../../intermediates/gradient/terrestrial/six_empos_fung_grad_data_ordered.rds")

bb <- lapply(all_bact_with_dummy, function(x) x[,sort(names(x))])
saveRDS(bb, "../../intermediates/gradient/terrestrial/six_empos_bact_grad_data_ordered.rds")
```









