---
title: "09_edaphic_factors"
output: html_document
date: "2023-06-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tibble)
library(dplyr)
library(igraph)
```

```{r}
# reformat soil data
edaphic_data <- read.csv("../data/cmaiki_soil_data.csv", header = FALSE)
edaphic_data <- edaphic_data[1:23,]

nums <- edaphic_data[3:nrow(edaphic_data), 3:ncol(edaphic_data)]


working_data <- nums
rownames(working_data) <- edaphic_data$V2[3:nrow(edaphic_data)]


colnames(working_data) <- c("CO2_burst_efflux",
                            "P_OLSEN",
                            "K",
                            "Ca",
                            "Mg",
                            "Na",
                            "pct_N",
                            "pct_C",
                            "pct_mega_WSA",
                            "pct_macro_WSA",
                            "water_holding_capacity",
                            "moisture_factor",
                            "pH",
                            "B_glucosidase",
                            "B_glucosiaminidase",
                            "PMN",
                            "HWEC",
                            "DOC:DON",
                            "soil_health_score",
                            "soil_health_category",
                            "substrate_bioavailibility",
                            "master_soil_properties",
                            "biological_activity",
                            "mineralogy_class",
                            "previously_intensive_ag_history")

# columns with no NAs in them
soil_data_complete <- working_data[,1:18]
soil_data_complete <- dplyr::mutate_all(soil_data_complete, as.numeric)

soil_data_complete$site_num <- substr(rownames(soil_data_complete), 6,6)

saveRDS(soil_data_complete, "../intermediates/gradient/soil_data_complete.rds")
```


```{r}

# import data
fung_data <- readRDS("~/cmaiki_lts/kaciekaj/waimea/intermediates/gradient/six_empos_fung_grad_data_ordered.rds")
bact_data <- readRDS("~/cmaiki_lts/kaciekaj/waimea/intermediates/gradient/six_empos_bact_grad_data_ordered.rds")

fung_meta_terr <- readRDS("../intermediates/gradient/fung_terrestrial_metadata.rds")
bact_meta_terr <- readRDS("../intermediates/gradient/bact_terrestrial_metadata.rds")

plots <- unique(fung_meta_terr$site_name)


remove_dummy_row <- function(df) {
  new_df <- df[row.names(df)!="dummy",]
  return(new_df)
}


fung_data <- lapply(fung_data, remove_dummy_row)
bact_data <- lapply(bact_data, remove_dummy_row)


# read in data = 20% prevalence global networks
fung_gradient_networks <- readRDS("../intermediates/gradient/fung_gradient_igraphs.rds")
bact_gradient_networks <- readRDS("../intermediates/gradient/bact_gradient_igraphs.rds")
cross_gradient_networks <- readRDS("../intermediates/gradient/cross_gradient_igraphs.rds")


```

Edaphic factors: mostly want to know can we say something additional about the ridge
Thinking PCA space - how network structure is explained or not by vectors of edaphic factors
Bray Curtis dissimilarity matrix

```{r}
# soil
## BETWEENNESS CENTRALITY
fung_bc <- sapply(fung_gradient_networks, betweenness)

for (i in 1:7) {
  names(fung_bc[[i]]) <- paste0("site", i, names(fung_bc[[i]]))
}

soil <- data.frame(btwn_cent = unlist(fung_bc))
soil$site_num <- substr(rownames(soil), 5,5)

edaphs <- names(soil_data_complete)[1:18]

for(i in 1:length(edaphs)) {
  soil[,i+2] <- soil_data_complete[,i][match(soil$site_num, soil_data_complete$site_num)]
  names(soil)[ncol(soil)] <- edaphs[i]
}




all_mrm <- MRM(dist(soil$btwn_cent) ~ dist(CO2_burst_efflux) + dist(`P_OLSEN`) + dist(K) + dist(Ca) + dist(Mg) + dist(Na) + dist(pct_N) + dist(pct_C) + dist(pct_mega_WSA) + dist(pct_macro_WSA) + dist(water_holding_capacity) + dist(moisture_factor) + dist(pH) + dist(B_glucosidase) + dist(B_glucosiaminidase) + dist(PMN) + dist(HWEC) + dist(`DOC:DON`), data = soil, nperm = 10)

saveRDS(all_mrm, "../intermediates/gradient/mrm_test.rds")

docdon_mrm <- MRM(dist(btwn_cent) ~ dist(docdon), data = soil, nperm = 10)

maybe <- MRM(dist(btwn_cent) ~ dist(docdon) + dist(water_holding_cap), data = soil, nperm = 10)




## by plot?

fung_diam <- sapply(fung_gradient_networks, diameter)

plot_df <- data.frame(diam = fung_diam)
plot_df$whc <- edaph_summ$water_holding_capacity

plot_whc_mrm <- whc_mrm <- MRM(dist(diam) ~ dist(whc), data = plot_df, nperm = 10)


### GLM?
idk_glm <- glm(btwn_cent ~CO2_burst_efflux + `P_OLSEN` + K + Ca + Mg + Na + pct_N + pct_C + pct_mega_WSA + pct_macro_WSA + water_holding_capacity + moisture_factor + pH + B_glucosidase + B_glucosiaminidase + PMN + HWEC + `DOC:DON`, data = soil)

model_2 = glm(proportion ~ Species * (Elev + Malaria), family = binomial, data = dat_no_iiwi_grped, weights = num)


```

???

```{r}
# another soil
wha <- capscale(dist(soil$btwn_cent) ~ water_holding_cap, data = soil, distance = "bray", add = TRUE)

capscale(formula = brazil.inv.and.non.spp ~ invaded + Condition(block), data = brazil.inv.and.non, distance = "bray", add = TRUE)
```




```{r}

soil_bc <- vegan::vegdist(soil$btwn_cent[soil$btwn_cent>0])
removed_rownames <- rownames(soil[soil$btwn_cent==0.000,])
kept_rownames <- rownames(soil[soil$btwn_cent>0,])

soil_pca <- prcomp(soil_bc, center = TRUE, scale= TRUE)

whc_bc <- vegan::vegdist(soil$water_holding_capacity[!(rownames(soil) %in% removed_rownames)])

whc_pca <- prcomp(whc_bc, center = TRUE, scale= TRUE)


bc_pca_dat <- data.frame(btwn_cent_pc1 = soil_pca$x[,1])
rownames(bc_pca_dat) <- kept_rownames

whc_pca_dat <- data.frame(whc_pc1 = whc_pca$x[,1])
rownames(whc_pca_dat) <- kept_rownames

test_pca_dat <- cbind(bc_pca_dat, whc_pca_dat$whc_pc1)
names(test_pca_dat)[2] <- "whc_pc1"

ord_pl <- ggplot(test_pca_dat, aes(x = btwn_cent_pc1, y = whc_pc1)) +
  geom_point() +
  #ggtitle("soil Gradient PC1 vs Soil pH") +
  #geom_smooth(method = "lm", se = FALSE) +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) 
  #scale_y_continuous(breaks = seq(8, 14, 2), limits = c(c(8, 14)))


## correlation test
y = cor.test(soil$btwn_cent, soil$water_holding_capacity, method = "spearman", exact=FALSE)

test_vec <- vector()
p_vec <- vector()
for (an_edaph in edaphs) {
  corr <- cor.test(soil$btwn_cent, soil[,an_edaph], method = "spearman", exact = FALSE)
  test_vec[an_edaph] <- ifelse(corr$p.value<=0.05, corr$estimate, 0)
  #p_vec[an_edaph] <- corr$p.value
  #ifelse(corr$p.value>0.05, test_vec[an_edaph]==0, test_vec[an_edaph])
}




# Variance explained by each component
pca_results <- summary(soil_pca)
ev <- soil_pca$sdev^2
newdat<-soil_pca$x[,1:4]


pca_dat <- as.data.frame(newdat)
#pca_dat$site_num <- fung_meta_terr$site_code[match(rownames(pca_dat), fung_meta_terr$sample_id)]

maybe_rownames <- rownames(soil[soil$btwn_cent>0,])
rownames(pca_dat) <- maybe_rownames

pca_dat$site_num <- substr(rownames(pca_dat), 5,5)

soil_data_complete$site_num <- substr(rownames(soil_data_complete), 6,6)

edaph_summ <- aggregate(. ~ site_num, data = soil_data_complete, FUN = mean)


pca_dat$pH <- edaph_summ$pH[match(pca_dat$site_num, edaph_summ$site_num)]




soil_lm <- lm(pca_dat$PC1 ~ pca_dat$pH)

soil_pl <- ggplot(pca_dat, aes(x = pH, y = PC1)) +
  geom_point() +
  ggtitle("soil Gradient PC1 vs Soil pH") +
  geom_smooth(method = "lm", se = FALSE) +
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) +
  #scale_y_continuous(breaks = seq(8, 14, 2), limits = c(c(8, 14))) +
  labs(subtitle = paste("Linear regression results:", "R2 =", round(summary(soil_lm)$r.squared, 3), "P =", round(summary(soil_lm)$coefficients[,4][2], 3)))


# Visualize the data points in the first two principal components
#plot(soil_pca$x[, 1], soil_pca$x[, 2])

# Access the loadings of each node on the first principal component
#loadings <- soil_pca$rotation[, 1]


plot(soil_pca, type="lines", main="eigenvalues")

```

```{r}
library(vegan)

### rows with only zeros cannot be handled by capscale
#spec = spec + 0.001

### select a dissimilarity index
rankindex(env,spec,indices = c("euc","bra"), method = "spearman")

### perform dbRDA
soil_dbRDA <- vegan::capscale(dist(soil$btwn_cent) ~ CO2_burst_efflux + `P_OLSEN` + K + Ca + Mg + Na + pct_N + pct_C + pct_mega_WSA + pct_macro_WSA + water_holding_capacity + moisture_factor + pH + B_glucosidase + B_glucosiaminidase + PMN + HWEC + `DOC:DON`, data = soil, dist="bray") 

ordiplot(soil_dbRDA,type="t",scaling=2)
ordiplot(soil_dbRDA,type="t",scaling=1)

### check constrained and unconstrained variance
soil_dbRDA$CCA$tot.chi/soil_dbRDA$tot.chi

### permutation tests to access significance of constraints
anova(soil_dbRDA) ## overall test of the significance 
anova(soil_dbRDA, by="terms", perm.max=999)

### correlation of the variables to the first two dbRDA axes
dbRDA.species.scores=scores(soil_dbRDA,choices =1:2,display = "sites")
cor(cbind(dbRDA.species.scores,soil[,3:ncol(soil)]))
```



Challenge: I can get PC1 from the Bray-Curtis matrix of co-occurrences, but each point is an OTU, and so I don't know how to regress this against the edaphic factors, which each correspond to a plot. A single OTU could be present in multiple plots.


## Chris's code

```{r}
# # Make environmental data as PC1
# ######## explore environmental data, use PCA to visualize
# ############### PCA
# 
# df.PCA<-environmental_data
# df.PCA$sampleID<-as.factor(rownames(df.PCA))
# df.PCA$HabitatType <- substr(df.PCA$sampleID, 0, 2) # extract first 2 letters of ID
# df.PCA$Plot <- substr(df.PCA$sampleID, 3, 3)
# df.PCA$Host <- substr(df.PCA$sampleID, 4, 5)
# df.PCA<-na.omit(df.PCA)
# df.PCA$HabitatType<-revalue(df.PCA$HabitatType, c("AK"="Restored Forest", "RO"="Remnant Forest"))
# 
# # remove columns unnecessary for final analysis, few factors retained
# env.PCA<-df.PCA[ , !names(df.PCA) %in% c("sampleID", "Plot", "Host", "HabitatType")]
# Hak.env.PCA <- prcomp(env.PCA, center = TRUE, scale= TRUE) # with HabitatType in dataframe
# PC.summary<-(summary(Hak.env.PCA))
# ev<-Hak.env.PCA$sdev^2
# newdat<-Hak.env.PCA$x[,1:4]
# # to see eigenvalues:  plot(Hak.env.PCA, type="lines", main="Hak.env.PCA eigenvalues")
# 
# 
# ############### PCA and NMDS combine
# NMDS.otu<-as.data.frame(NMDS1) # NMDS Bray Curtis of OTUs
# PCA.env<-as.data.frame(newdat) # the exported data from PCA of environment
# PC.NMD<-merge(PCA.env, NMDS.otu, by = "row.names", all = TRUE) # merge dataframes
# PC.NMD<-na.omit(PC.NMD) #drop NA columns that don't correspond
# colnames(PC.NMD)[1]<-"sampleID"
# PC.NMD$HabitatType <- as.factor(substr(PC.NMD$sampleID, 0, 2)) # extract first 2 letters of ID
# PC.NMD$HabitatType<-revalue(PC.NMD$HabitatType, c("AK"="Restored Forest", "RO"="Remnant Forest"))
# PC.NMD$HabitatType<- factor(PC.NMD$HabitatType, levels=c("Remnant Forest", "Restored Forest"))
# PC.NMD$Plot <- as.factor(substr(PC.NMD$sampleID, 3, 3)) # make plot ID
# PC.NMD$Host <- as.factor(substr(PC.NMD$sampleID, 4, 5)) # make host ID
# 
# 
# ##### ##### ##### ##### ##### ##### 
# ## soils for relationship pf PC x NMDS
# 
# # full model
# mod<-lm(NMDS1~PC1*HabitatType, data=PC.NMD)
# print(anova(mod), digits=6)
# 
# # soil RO/Remnant Forest relationship
# RO.PCNMD<-PC.NMD[(PC.NMD$HabitatType=="Remnant Forest"),]
# mod.RO<-lm(NMDS1~PC1, data=RO.PCNMD); print(anova(mod.RO), digits=6)
# 
# # soil AK/Restored Forest relationship
# AK.PCNMD<-PC.NMD[(PC.NMD$HabitatType=="Restored Forest"),]
# mod.AK<-lm(NMDS1~PC1, data=AK.PCNMD); print(anova(mod.AK), digits=6)
# 
# 
# ### formatting
# BW.back<-theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#                panel.background = element_blank(), axis.line = element_line(colour = "black", size=0.2))
# 
# ### ### ### ### ### ### 
# ### plot of NMDS-1 x PC1
# PC.NMDS.plot<-ggplot(data=PC.NMD, aes(x=PC1, y=NMDS1, color=HabitatType))+
#   geom_point(aes(color=HabitatType), alpha=.7) + 
#   scale_color_manual(values=Hab.col) + 
#   scale_fill_manual(values=Hab.col) +
#   xlab("PC1 (soil chemistry, 43% variance explained)") +
#   ylab("NMDS1 (AMF Bray-Curtis)") +
#   theme_bw()+
#   geom_smooth(aes(color=HabitatType, fill=HabitatType), values=Hab.col, method = "lm", alpha = .2, size=0.6) +
#   theme(legend.position="top", legend.title = element_blank(),
#         axis.text=element_text(size=6),
#         axis.title=element_text(size=10)) +
#   BW.back
# 
# PC.NMDS.plot
# dev.copy(pdf, "figures/execute code/PC.NMD.slopes.pdf", height=3, width=6)
# dev.off()
```


sandbox

```{r}
library(vegan)
trees <- read.csv("https://raw.githubusercontent.com/pbuttigieg/marmicStats/master/statsII/trees.csv",header=T)

#### prepare the data
row.names(trees)<-trees$ECOSYS
spec<-trees[,11:23]
env<- trees[,3:10]

### rows with only zeros cannot be handled by capscale
spec = spec + 0.001

### perform dbRDA
mydbRDA <- vegan::capscale(spec ~ var1 + var2 + var3 + var4, data = env, dist="bray") 

ordiplot(mydbRDA,type="t",scaling=2)
ordiplot(mydbRDA,type="t",scaling=1)

### check constrained and unconstrained variance
mydbRDA$CCA$tot.chi/mydbRDA$tot.chi

### permutation soils to access significance of constraints
anova(mydbRDA) ## overall soil of the significance 
anova(mydbRDA, by="terms", perm.max=999)

### correlation of the variables to the first two dbRDA axes
dbRDA.species.scores=scores(mydbRDA,choices =1:2,display = "sites")
cor(cbind(dbRDA.species.scores,env[,1:4]))
```

# tutorial?
```{r}
ab_trees = read.csv('../misc/AB_Climate_Trees.csv')
rownames(ab_trees) = ab_trees$ECOSYS # creates rownames

ecolabels = as.factor(ab_trees$ECOSYS) # creates labels we need later
trees = ab_trees[,11:23]
climate = ab_trees[,3:10]
trees = trees + 0.000001 #adjust for distance matrixes
head(trees) # check it
head(climate) # check it
scaled_clim = scale(climate)

rda = rda(trees ~ MAT+MWMT+MCMT+TD+lnMAP+lnMSP+lnAHM+lnSHM, climate)
summary(rda)

## dbrda
euc = vegdist(trees, 'euclidean')
dbRDA = capscale(euc ~ MAT+MWMT+MCMT+TD+lnMAP+lnMSP+lnAHM+lnSHM,
climate)
summary(dbRDA)

scores2 = data.frame(CAP1 = dbRDA$CCA$wa[,1], CAP2 = dbRDA$CCA$wa[,2])
plot(scores2, col='black', xlim=c(-1,1))
text(scores2, labels=row.names(scores2))
vectors3=envfit(scores2, scaled_clim, nperm=0)
plot(vectors3, col="blue")
vectors4=envfit(scores2, trees, nperm=0)
plot(vectors4, col=c('red'))

plot(dbRDA)

```


```{r}
library(ecodist)
data(graze)
# Abundance of this grass is related to forest cover but not location
MRM(dist(LOAR10) ~ dist(sitelocation) + dist(forestpct), data=graze, nperm=10)
# Abundance of this legume is related to location but not forest cover
MRM(dist(TRRE3) ~ dist(sitelocation) + dist(forestpct), data=graze, nperm=10)
# Compare to presence/absence of grass LOAR10 using logistic regression
LOAR10.presence <- ifelse(graze$LOAR10 > 0, 1, 0)
MRM(dist(LOAR10.presence) ~ dist(sitelocation) + dist(forestpct),
data=graze, nperm=10, method="logistic")
```



