---
title: "09c_edaphic_pc_regress"
output: html_document
date: "2023-07-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
soil_data_complete <- readRDS("../intermediates/gradient/soil_data_complete.rds")
edaph_summ <- aggregate(. ~ site_num, data = soil_data_complete, FUN = mean)
edaphs <- names(soil_data_complete)[1:18]
```

```{r}

# import data
fung_data <- readRDS("~/cmaiki_lts/kaciekaj/waimea/intermediates/gradient/six_empos_fung_grad_data_ordered.rds")
bact_data <- readRDS("~/cmaiki_lts/kaciekaj/waimea/intermediates/gradient/six_empos_bact_grad_data_ordered.rds")

fung_meta_terr <- readRDS("../intermediates/gradient/fung_terrestrial_metadata.rds")
bact_meta_terr <- readRDS("../intermediates/gradient/bact_terrestrial_metadata.rds")
cross_meta_terr <- plyr::rbind.fill(fung_meta_terr, bact_meta_terr)

plots <- unique(fung_meta_terr$site_name)


remove_dummy_row <- function(df) {
  new_df <- df[row.names(df)!="dummy",]
  return(new_df)
}


fung_data <- lapply(fung_data, remove_dummy_row)
bact_data <- lapply(bact_data, remove_dummy_row)


# read in data = 20% prevalence global networks
fung_gradient_networks <- readRDS("../intermediates/gradient/fung_gradient_igraphs.rds")
bact_gradient_networks <- readRDS("../intermediates/gradient/bact_gradient_igraphs.rds")
cross_gradient_networks <- readRDS("../intermediates/gradient/cross_gradient_igraphs.rds")


```

```{r, message=FALSE}
library(tibble)
library(dplyr)
library(igraph)
library(ggplot2)
```


Goal 1: Cut down soil factors to the meaningful ones, run ANOVAs to check that they do change significantly across the gradient
Goal 2: Generate a Bray-Curtis distance matrix from network metrics and PD by sample. Regress the PC1 of a PCA on this matrix against the log-fold change in edaphic factor across the gradient.

### ANOVAs to weed out nonsignificant changes

```{r}
soil_aov_pvals <- vector()

for (an_edaph in edaphs) {
  my_lm <- lm(soil_data_complete[,an_edaph] ~ soil_data_complete$site_num)
  my_anova <- aov(my_lm)
  
  soil_aov_pvals[an_edaph] <- summary(my_anova)[[1]][["Pr(>F)"]][1]
}

# all are significant except pct_N, pct_C, PMN
```

```{r}
# filter edaphic data to more relevant factors
soil_data_filtered <- select(soil_data_complete, site_num, CO2_burst_efflux, `P_OLSEN`, K, Ca, Mg, HWEC, water_holding_capacity, pH)

edaph_summ_filtered <- edaph_summ[,colnames(edaph_summ) %in% colnames(soil_data_filtered)]
```


### MPD Euclidean Matrix

```{r}
mpd_sample_data <- readRDS("../intermediates/gradient/samp_phydist_data.rds")

mpd_dist_matrix <- dist(mpd_sample_data$mpd.obs.z)

# on observed mpd, not z scores
#mpd_bc_matrix <- vegan::vegdist(mpd_sample_data$mpd.obs)

mpd_pca <- prcomp(mpd_dist_matrix, center = TRUE, scale = TRUE)

# pc1and2 <- data.frame(mpd_pca$x[,1:2])
# rownames(pc1and2) <- rownames(mpd_sample_data)
# pc1and2$site_num <- cross_meta_terr$site_code[match(rownames(pc1and2), cross_meta_terr$sample_id)]

mpd_pca_dat <- data.frame(mpd_pca$x[,1:2])
rownames(mpd_pca_dat) <- rownames(mpd_sample_data)
mpd_pca_dat$site_num <- cross_meta_terr$site_code[match(rownames(mpd_pca_dat), cross_meta_terr$sample_id)]

filtered_edaphs <- names(soil_data_filtered[2:ncol(soil_data_filtered)])

for (i in 1:length(filtered_edaphs)) {
  mpd_pca_dat[,i+3] <- edaph_summ_filtered[,i+1][match(mpd_pca_dat$site_num, edaph_summ_filtered$site_num)]
  names(mpd_pca_dat)[ncol(mpd_pca_dat)] <- filtered_edaphs[i]
}


mpd_pca_dat$site_name <- cross_meta_terr$site_name[match(mpd_pca_dat$site_num, cross_meta_terr$site_code)]
mpd_pca_dat$site_name <- factor(mpd_pca_dat$site_name, levels=c("Beach", "Estuary", "Entrance", "Confluence", "Waterfall", "DrumRoad", "Ridge"))


## log fold change?
# test_vals <- edaph_summ_filtered$CO2_burst_efflux
# names(test_vals) <- edaph_summ_filtered$site_num
# test_vals <- sort(test_vals)
# 
# lfc1 <- function(a, b) log((b-a)/a + 1)
# 
# lfc1(edaph_summ_filtered[1,2], edaph_summ_filtered[2,2])
# 
# lfc_res <- vector()
# for (i in 1:7) {
#   first <- test_vals[1]
#   second <- test_vals[i]
#   
#   lfc_res[i] <- lfc1(first, second) 
# }
# names(lfc_res) <- names(test_vals)
# 
# mpd_pca_dat$co2_lfc_recode <- lfc_res[match(mpd_pca_dat$site_num, names(lfc_res))]


mpd_edaph_plots <- list()

for (an_edaph in filtered_edaphs) {
  my_lm <- lm(mpd_pca_dat[,"PC1"] ~ mpd_pca_dat[,an_edaph])
  
  my_plot <- ggplot(mpd_pca_dat, aes_string(x = an_edaph, y = "PC1", color = "site_name")) +
    geom_point() +
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) +
  labs(title= an_edaph, x = an_edaph, subtitle = paste("Linear regression:", "R2 =", round(summary(my_lm)$r.squared, 3), ";", "P =", round(summary(my_lm)$coefficients[,4][2], 3)))

  mpd_edaph_plots[[an_edaph]] <- my_plot
  
}

library(ggpubr)

all_mpd_edaph <- ggarrange(plotlist=mpd_edaph_plots, ncol = 2, nrow = 4, common.legend = TRUE, legend = "right")

ggsave("../figures/gradient/mpd_edaphic_regress_plots.png", all_mpd_edaph, width = 15, height = 25)

```


### Network Metric Bray-Curtis PCAs

Function for making subgraphs by sample

```{r}
make_sample_subnets <- function(abun, network) {
  results <- list()
  for (i in 1:ncol(abun)) {
  to_sub <- rownames(abun[abun[i]>0,])
  subnet <- induced_subgraph(network, V(network)$name %in% to_sub)
  
  results[[i]] <- subnet
  names(results)[i] <- names(abun)[i]
  }
  return(results)
}

```

### ITS

```{r}
fung_sample_subnets <- mapply(make_sample_subnets, fung_data, fung_gradient_networks, SIMPLIFY=FALSE)


# diameter
fung_diam <- lapply(fung_sample_subnets, function(x) lapply(x, diameter))

fung_diam_dat <- data.frame(diam = unlist(fung_diam))
fung_diam_dat$site_num <- fung_meta_terr$site_code[match(rownames(fung_diam_dat), fung_meta_terr$sample_id)]

for(i in 1:length(edaphs)) {
  fung_diam_dat[,i+2] <- edaph_summ[,i+1][match(fung_diam_dat$site_num, edaph_summ$site_num)]
  names(fung_diam_dat)[ncol(fung_diam_dat)] <- edaphs[i]
}


fung_diam_dist_matrix <- vegan::vegdist(fung_diam_dat$diam[fung_diam_dat$diam>0])

fung_diam_pca <- prcomp(fung_diam_dist_matrix, center = TRUE, scale = TRUE)

# pc1and2 <- data.frame(fung_diam_pca$x[,1:2])
# rownames(pc1and2) <- rownames(fung_diam_sample_data)
# pc1and2$site_num <- cross_meta_terr$site_code[match(rownames(pc1and2), cross_meta_terr$sample_id)]

fung_diam_pca_dat <- data.frame(fung_diam_pca$x[,1:2])
rownames(fung_diam_pca_dat) <- rownames(fung_diam_dat[fung_diam_dat$diam>0,])
fung_diam_pca_dat$site_num <- cross_meta_terr$site_code[match(rownames(fung_diam_pca_dat), cross_meta_terr$sample_id)]

for (i in 1:length(filtered_edaphs)) {
  fung_diam_pca_dat[,i+3] <- edaph_summ_filtered[,i+1][match(fung_diam_pca_dat$site_num, edaph_summ_filtered$site_num)]
  names(fung_diam_pca_dat)[ncol(fung_diam_pca_dat)] <- filtered_edaphs[i]
}


fung_diam_pca_dat$site_name <- cross_meta_terr$site_name[match(fung_diam_pca_dat$site_num, cross_meta_terr$site_code)]
fung_diam_pca_dat$site_name <- factor(fung_diam_pca_dat$site_name, levels=c("Beach", "Estuary", "Entrance", "Confluence", "Waterfall", "DrumRoad", "Ridge"))



fung_diam_edaph_plots <- list()

for (an_edaph in filtered_edaphs) {
  my_lm <- lm(fung_diam_pca_dat[,"PC1"] ~ fung_diam_pca_dat[,an_edaph])
  
  my_plot <- ggplot(fung_diam_pca_dat, aes_string(x = an_edaph, y = "PC1", color = "site_name")) +
    geom_point() +
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) +
  labs(title= an_edaph, x = an_edaph, subtitle = paste("Linear regression:", "R2 =", round(summary(my_lm)$r.squared, 3), ";", "P =", round(summary(my_lm)$coefficients[,4][2], 3)))

  fung_diam_edaph_plots[[an_edaph]] <- my_plot
  
}

all_fung_diam_edaph <- ggarrange(plotlist=fung_diam_edaph_plots, ncol = 2, nrow = 4, common.legend = TRUE, legend = "right")

ggsave("../figures/gradient/fung_diam_edaphic_regress_plots_bray_curtis.png", all_fung_diam_edaph, width = 15, height = 25)

```

### 16S

```{r}
bact_sample_subnets <- mapply(make_sample_subnets, bact_data, bact_gradient_networks, SIMPLIFY=FALSE)


# diameter
bact_diam <- lapply(bact_sample_subnets, function(x) lapply(x, diameter))

bact_diam_dat <- data.frame(diam = unlist(bact_diam))
bact_diam_dat$site_num <- bact_meta_terr$site_code[match(rownames(bact_diam_dat), bact_meta_terr$sample_id)]

for(i in 1:length(edaphs)) {
  bact_diam_dat[,i+2] <- edaph_summ[,i+1][match(bact_diam_dat$site_num, edaph_summ$site_num)]
  names(bact_diam_dat)[ncol(bact_diam_dat)] <- edaphs[i]
}


bact_diam_dist_matrix <- vegan::vegdist(bact_diam_dat$diam)

bact_diam_pca <- prcomp(bact_diam_dist_matrix, center = TRUE, scale = TRUE)

# pc1and2 <- data.frame(bact_diam_pca$x[,1:2])
# rownames(pc1and2) <- rownames(bact_diam_sample_data)
# pc1and2$site_num <- cross_meta_terr$site_code[match(rownames(pc1and2), cross_meta_terr$sample_id)]

bact_diam_pca_dat <- data.frame(bact_diam_pca$x[,1:2])
rownames(bact_diam_pca_dat) <- rownames(bact_diam_dat)
bact_diam_pca_dat$site_num <- cross_meta_terr$site_code[match(rownames(bact_diam_pca_dat), cross_meta_terr$sample_id)]

for (i in 1:length(filtered_edaphs)) {
  bact_diam_pca_dat[,i+3] <- edaph_summ_filtered[,i+1][match(bact_diam_pca_dat$site_num, edaph_summ_filtered$site_num)]
  names(bact_diam_pca_dat)[ncol(bact_diam_pca_dat)] <- filtered_edaphs[i]
}


bact_diam_pca_dat$site_name <- cross_meta_terr$site_name[match(bact_diam_pca_dat$site_num, cross_meta_terr$site_code)]
bact_diam_pca_dat$site_name <- factor(bact_diam_pca_dat$site_name, levels=c("Beach", "Estuary", "Entrance", "Confluence", "Waterfall", "DrumRoad", "Ridge"))



bact_diam_edaph_plots <- list()

for (an_edaph in filtered_edaphs) {
  my_lm <- lm(bact_diam_pca_dat[,"PC1"] ~ bact_diam_pca_dat[,an_edaph])
  
  my_plot <- ggplot(bact_diam_pca_dat, aes_string(x = an_edaph, y = "PC1", color = "site_name")) +
    geom_point() +
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) +
  labs(title= an_edaph, x = an_edaph, subtitle = paste("Linear regression:", "R2 =", round(summary(my_lm)$r.squared, 3), ";", "P =", round(summary(my_lm)$coefficients[,4][2], 3)))

  bact_diam_edaph_plots[[an_edaph]] <- my_plot
  
}

all_bact_diam_edaph <- ggarrange(plotlist=bact_diam_edaph_plots, ncol = 2, nrow = 4, common.legend = TRUE, legend = "right")

ggsave("../figures/gradient/bact_diam_edaphic_regress_plots_bray_curtis.png", all_bact_diam_edaph, width = 15, height = 25)

```


### Cross-domain

```{r}
cross_data <- list()

for (i in 1:length(fung_data)) {
  cross <- plyr::rbind.fill(fung_data[[i]], bact_data[[i]])
  rownames(cross) <- c(rownames(fung_data[[i]]), rownames(bact_data[[i]]))
  cross_data[[i]] <- cross
}

saveRDS(cross_data, "../intermediates/gradient/cross_data.rds")

cross_meta_terr <- plyr::rbind.fill(fung_meta_terr, bact_meta_terr)
saveRDS(cross_meta_terr, "../intermediates/gradient/cross_metadata_terr.rds")

cross_sample_subnets <- mapply(make_sample_subnets, cross_data, cross_gradient_networks, SIMPLIFY=FALSE)


# diameter
cross_diam <- lapply(cross_sample_subnets, function(x) lapply(x, diameter))

cross_diam_dat <- data.frame(diam = unlist(cross_diam))
cross_diam_dat$site_num <- cross_meta_terr$site_code[match(rownames(cross_diam_dat), cross_meta_terr$sample_id)]

for(i in 1:length(edaphs)) {
  cross_diam_dat[,i+2] <- edaph_summ[,i+1][match(cross_diam_dat$site_num, edaph_summ$site_num)]
  names(cross_diam_dat)[ncol(cross_diam_dat)] <- edaphs[i]
}


cross_diam_dist_matrix <- vegan::vegdist(cross_diam_dat$diam)

cross_diam_pca <- prcomp(cross_diam_dist_matrix, center = TRUE, scale = TRUE)

cross_diam_pca_dat <- data.frame(cross_diam_pca$x[,1:2])
rownames(cross_diam_pca_dat) <- rownames(cross_diam_dat)
cross_diam_pca_dat$site_num <- cross_meta_terr$site_code[match(rownames(cross_diam_pca_dat), cross_meta_terr$sample_id)]

for (i in 1:length(filtered_edaphs)) {
  cross_diam_pca_dat[,i+3] <- edaph_summ_filtered[,i+1][match(cross_diam_pca_dat$site_num, edaph_summ_filtered$site_num)]
  names(cross_diam_pca_dat)[ncol(cross_diam_pca_dat)] <- filtered_edaphs[i]
}


cross_diam_pca_dat$site_name <- cross_meta_terr$site_name[match(cross_diam_pca_dat$site_num, cross_meta_terr$site_code)]
cross_diam_pca_dat$site_name <- factor(cross_diam_pca_dat$site_name, levels=c("Beach", "Estuary", "Entrance", "Confluence", "Waterfall", "DrumRoad", "Ridge"))



cross_diam_edaph_plots <- list()

for (an_edaph in filtered_edaphs) {
  my_lm <- lm(cross_diam_pca_dat[,"PC1"] ~ cross_diam_pca_dat[,an_edaph])
  
  my_plot <- ggplot(cross_diam_pca_dat, aes_string(x = an_edaph, y = "PC1", color = "site_name")) +
    geom_point() +
    theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"),
        panel.background=element_blank()) +
  labs(title= an_edaph, x = an_edaph, subtitle = paste("Linear regression:", "R2 =", round(summary(my_lm)$r.squared, 3), ";", "P =", round(summary(my_lm)$coefficients[,4][2], 3)))

  cross_diam_edaph_plots[[an_edaph]] <- my_plot
  
}

all_cross_diam_edaph <- ggarrange(plotlist=cross_diam_edaph_plots, ncol = 2, nrow = 4, common.legend = TRUE, legend = "right")

ggsave("../figures/gradient/cross_diam_edaphic_regress_plots_bray_curtis.png", all_cross_diam_edaph, width = 15, height = 25)

```


```{r}
saveRDS(list(fung_diam_dat, bact_diam_dat, cross_diam_dat), "../intermediates/gradient/all_diameter_data_for_mrm.rds")
```




