---
title: "8_abundance_occupancy"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tibble)
library(reshape2)
```

```{r}
## Source cdmuir header ----
source("../abundance-occupancy-main/header.R")
```


```{r}
old_data <- read.csv("../../waimea/abundance-occupancy-main/data_for_regress.csv")

old_dat <- old_data %>%
  mutate(
    log_asv_abundance = log(asv_abundance),
    log_sample_count = log(sample_count),
    group = case_when(
      soil_host_core_color == "black" & host_spec_color == "black" ~ 1,
      soil_host_core_color == "darkgoldenrod1" & host_spec_color == "black" ~ 2,
      soil_host_core_color == "black" & host_spec_color == "slateblue1" ~ 3,
      soil_host_core_color == "darkgoldenrod1" & host_spec_color == "slateblue1" ~ 4,
    )
  )
```

The goal here is to apply Chris Muir's abundance-occupancy code/model to our Waimea data, but it's going to take a bit for me to wrap my head around how to translate since I was not a very good coder back when he used my data to make this. Plus, the Waimea data has a different structure than Hakalau.

The data frame sent to him had 5 colummns: ASV name, sample_count, asv_abundance, and columns that indicated whether the ASV was core or host-specific. The problem is the identity was indicated using colors, since this was the data I used to plot the regression scatterplot. So column soil_host_core_color has values of either "black" or "darkgoldenrod1" and host_spec_color is either "black" or "slateblue1". 

```{r}
old_data
```

I looked in the model code (model.stan) and it doesn't seem like the color names are used at all. 
However the colors are somewhat involved in the linear model used to make matrix x as seen below.

```{r}
#old_dat <- old_dat[sample(nrow(old_dat), 100),]

old_dat_list <- list(
  x = model.matrix(lm(log_asv_abundance ~ soil_host_core_color *
                        host_spec_color,  data = old_dat)),
  y = as.matrix(select(dat, log_asv_abundance, log_sample_count))
  )

old_dat_list$y1 <- old_dat$asv_abundance
old_dat_list$y2 <- old_dat$sample_count

old_dat_list$n_obs <- nrow(old_dat_list$x)

old_dat_list$J <- ncol(old_dat_list$x)
old_dat_list$K <- ncol(old_dat_list$y)
```

Okay if I change the color names to yeses and nos (strings), the model numbers still come out the same.

Hm, but in the model.. Intercept is core color black and host spec color black (neither core nor host spec)
Then the next coefficient is for core color goldenrod (is core)
Then it's host spec color slateblue (is host specific)
Then it's the empty one with both goldenrod and slateblue, for which no ASVs occur.

Okay. So the two variables in the model formul give rise to our four categories (others, core, host spec, both).

To keep things simpler, I don't want to mess with this formula, and I'll need to code the "core" and "host spec" equivalent columns in the Waimea version so that our desired categories are honored. 

"yes" "yes" should be a trophic level (along with core and host spec), and then "no" "no" should be left as the residuals (others). 

Maybe the contents of these columns alone aren't readable per se, but another column will identify (for me) which trophic level an ASV belongs to. 

I'll run it once per habitat (terrestrial, riverine, marine) and the levels will be others, primary producer, environmental, and consumer.

"yes" and "no" will be okay because of the alphabet order, intercept will still be no no.


```{r}
# waimea version
meta <- readRDS("../intermediates/culled_metadata_edited_trophic.rds")
abun <- readRDS("../intermediates/culled_asv_table.rds")

```

Steps (terrestrial example):
Subset metadata table where habitat == terrestrial
Subset asv table where sample ids match between meadata and asv table (remove empty ASVs)

OR
Follow what I did in 7_core.rmd, melt the whole ASV table, then use match to bring in habitat and trophic level data based on sample
Then subset by habitat
group by ASV, summarise with n_distinct to get number of samples it's in

Based on trophic level, make columns to use in lm()

consumer, environment, primary producer are trophic levels in alpha order

```{r}

abun2 <- rownames_to_column(abun, "ASV")
melt_abun <- melt(abun2)
melt_abun <- melt_abun[melt_abun$value>0,]
colnames(melt_abun)[2] <- "sequencing_id"
colnames(melt_abun)[3] <- "asv_abundance"

# add habitat and trophic columns matching by sequencing ID
melt_abun$habitat <- meta$habitat[match(melt_abun$sequencing_id, meta$sequencing_id)]
melt_abun$trophic <- meta$trophic[match(melt_abun$sequencing_id, meta$sequencing_id)]

trophs = c("Consumer", "Environmental", "PrimaryProducer")
habitats = c("Terrestrial", "Riverine", "Marine")


hab_dfs <- list()

for (a_hab in habitats) {
  sub_df <- melt_abun[which(melt_abun$habitat==a_hab),]
  hab_dfs[[a_hab]] <- sub_df
}

samp_count_calc <- function(df) {
  # df will already be subset by habitat
  df <- as.data.frame(df)
  df <- df %>% group_by(ASV) %>% summarise(sample_count = n_distinct(sequencing_id))
  return(df)
}

a <- lapply(hab_dfs, samp_count_calc)

# match sample count to make new column in habitat dfs
for (a_hab in habitats) {
  df <- hab_dfs[[a_hab]]
  ct <- a[[a_hab]]
  df$sample_count <- ct$sample_count[match(df$ASV, ct$ASV)]
  hab_dfs[[a_hab]] <- df
}

test_df <- hab_dfs[["Marine"]]

# 3 columns (mod_col1, mod_col2)
# 
# where trophic = consumer, col1 = "yes", col2 = "no"
# where trophic = environment, col1 = "no", col2 = "yes
# where trophic = primary producer, col1 = "yes", col2 = "yes"
for (a_hab in habitats) {
  df <- hab_dfs[[a_hab]]
  df <- df %>% mutate(log_asv_abundance = log(asv_abundance),
                      log_sample_count = log(sample_count),
                      mod_col1 = case_when(trophic=="Consumer" ~ "yes",
                                        trophic=="Environmental" ~ "no",
                                        trophic=="PrimaryProducer" ~ "yes"),
                      mod_col2 = case_when(trophic=="Consumer" ~ "no",
                                        trophic=="Environmental" ~ "yes",
                                        trophic=="PrimaryProducer" ~ "yes"),
                      group = case_when(trophic=="Consumer" ~ 1,
                                        trophic=="Environmental" ~ 2,
                                        trophic=="PrimaryProducer" ~ 3))
  hab_dfs[[a_hab]] <- df
}

saveRDS(hab_dfs, "../intermediates/abun_occ_waimea_habitat_dfs.rds")
```


```{r}
## Convert data to list for Stan ----

# 1st try with riverine
dat = hab_dfs[["Riverine"]]

dat <- dat[sample(nrow(dat), 100), ]

str_dat_list <- list(
  x = model.matrix(lm(log_asv_abundance ~ mod_col1 * mod_col2,  data = dat)),
  y = as.matrix(select(dat, log_asv_abundance, log_sample_count))
  )

str_dat_list$y1 <- dat$asv_abundance
str_dat_list$y2 <- dat$sample_count

str_dat_list$n_obs <- nrow(str_dat_list$x)

str_dat_list$J <- ncol(str_dat_list$x)
str_dat_list$K <- ncol(str_dat_list$y)
```


```{r}
## Compile Stan model (this only needs to be done once) ----
mod <- cmdstan_model("../abundance-occupancy-main/model.stan")

## Fit model ----
str_fit <- mod$sample(
  data = str_dat_list,
  seed = 89638177,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)

## Fit model ----
old_fit <- mod$sample(
  data = old_dat_list,
  seed = 89638177,
  chains = 4,
  parallel_chains = 4,
  refresh = 500
)

## Save fitted model ----
fit$save_object(file = "fit.rds")
write_rds(fit$metadata(), "fit-metadata.rds")
```

Error

grep: --color=never: No such file or directory
NAs introduced by coercionNAs introduced by coercionNAs introduced by coercionNAs introduced by coercionError in if (inv_metric_rows > 0 && csv_file_info$metric == "dense_e") { : 
  missing value where TRUE/FALSE needed
  
```{r}
##  Make a figure of model predictions with data

# Make data for plotting ----
df <- fit$draws("beta") %>%
  as_draws_df() %>%
  mutate(
    log_asv_abundance_1 = `beta[1,1]`,
    log_sample_count_1 = `beta[2,1]`,
    log_asv_abundance_2 = `beta[1,1]` + `beta[1,2]`,
    log_sample_count_2 = `beta[2,1]` + `beta[2,2]`,
    log_asv_abundance_3 = `beta[1,1]` + `beta[1,3]`,
    log_sample_count_3 = `beta[2,1]` + `beta[2,3]`,
    log_asv_abundance_4 = `beta[1,1]` + `beta[1,2]` + `beta[1,3]` + `beta[1,4]`,
    log_sample_count_4 = `beta[2,1]` + `beta[2,2]` + `beta[2,3]` + `beta[2,4]`,
  ) %>%
  select(starts_with("log"))

# Hypothesis testing
hypothesis(df, "log_asv_abundance_2 = log_asv_abundance_3")

df_mut <- df %>%
  pivot_longer(everything(), values_to = "estimate") %>%
  group_by(name) %>%
  point_interval() %>%
  mutate(
    group = str_extract(name, "[1-4]{1}$"),
    name = str_remove(name, "_[1-4]{1}$")
  ) %>%
  pivot_wider(values_from = c(estimate, .lower, .upper)) %>%
  # only including core and host-specific ASVs
  filter(group %in% c(2,3))

# make a column in the data df with only 2 color options
# aka removing instances of "black" 
dat_no_others = dat[
  (dat$soil_host_core_color == "darkgoldenrod1") | 
    (dat$host_spec_color == "slateblue1"),]

dat_no_others$plot_color = dat_no_others$soil_host_core_color
dat_no_others$plot_color = str_replace(dat_no_others$plot_color, 
                                       "black", "slateblue1")


## Make figure ----
ggplot(df_mut, aes(
  x = estimate_log_sample_count, y = estimate_log_asv_abundance,
  xmin = .lower_log_sample_count, ymin = .lower_log_asv_abundance,
  xmax = .upper_log_sample_count, ymax = .upper_log_asv_abundance,
  color = as.factor(group)
)) +
  facet_grid(group ~ ., labeller = labeller(
    group = c(
      #`1` = "Others",
      `2` = "Core Host + Soil",
      `3` = "Host-specific"
      #`4` = "both"
    ))) +
  geom_point(
    data = dat_no_others,
    mapping = aes(x = log_sample_count, y = log_asv_abundance,
                  color = as.factor(group)), 
    inherit.aes = FALSE, alpha = 0.2
  ) +
  geom_linerange(orientation = "x", size = 2) +
  geom_linerange(orientation = "y", size = 2) +
  geom_point(size = 5, shape = 21, fill = "white") +
  scale_color_manual(values = c("darkgoldenrod1", "slateblue1")) +
  xlab("log(number_of_samples)") +
  ylab("log(asv_abundance)") +
  theme_cowplot() +
  theme(legend.position = "none")
```



