---
title: "11_funguild_checks"
output: html_document
date: "2024-04-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE, echo=FALSE}
library(igraph)
```

Goal: look at the fungi taking part in negative interactions (check interkingdom) and see if they are more likely to be parasitic or pathogens in funguild

```{r, echo = FALSE, results = "hide"}
filt_cross_net <- readRDS("../intermediates/global/weighted_filt_cross_net.rds")

filt_cross_net <- delete_vertices(filt_cross_net, "dummy")

filt_bact_net <- readRDS("../intermediates/global/weighted_filt_bact_net.rds")

filt_bact_net <- delete_vertices(filt_bact_net, "dummy")

fung_abun <- readRDS("../intermediates/global/downsampled_fung_otu_table.rds")
bact_abun <- readRDS("../intermediates/global/downsampled_bact_otu_table.rds")

fung_abun <- fung_abun[rownames(fung_abun)!="dummy",]
bact_abun <- bact_abun[rownames(bact_abun)!="dummy",]


funguilds <- readRDS("../intermediates/gradient/terrestrial/funguild_assignments.rds")
```

```{r, echo=FALSE, results = "hide"}
# code chunk from 05a_downsample_modularity.Rmd
filt_cross_net

x = igraph::as_data_frame(filt_cross_net)[, ]

fung_bact <- x[substr(x$from, 1, 3)!=substr(x$to, 1, 3),]

fb_neg <- fung_bact[fung_bact$sign=="negative",]

fb_fung_neg <- unique(fb_neg$from)

fbn <- data.frame(my_otu = fb_fung_neg)

fbn$taxonomy <- funguilds$Taxonomy[match(fbn$my_otu, funguilds$OTU)]
fbn$taxlevel <- funguilds$taxon[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_trophicMode <- funguilds$trophicMode[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_guild <- funguilds$guild[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_conf <- funguilds$confidenceRanking[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_growthform <- funguilds$growthForm[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_trait <- funguilds$trait[match(fbn$my_otu, funguilds$OTU)]
fbn$fung_notes <- funguilds$notes[match(fbn$my_otu, funguilds$OTU)]

fb_neg_assigned <- fbn[!is.na(fbn$taxlevel),]

#table(fb_neg_assigned$fung_trophicMode)/nrow(fb_neg_assigned)
table(fb_neg_assigned$fung_trophicMode)/length(fb_fung_neg)

# what proportion of fungal nodes could be assigned guild
dim(fb_neg_assigned)[1] / dim(fbn)[1]


fb_pos <- fung_bact[fung_bact$sign=="positive",]

fb_fung_pos <- unique(fb_pos$from)

fbp <- data.frame(my_otu = fb_fung_pos)

fbp$taxonomy <- funguilds$Taxonomy[match(fbp$my_otu, funguilds$OTU)]
fbp$taxlevel <- funguilds$taxon[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_trophicMode <- funguilds$trophicMode[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_guild <- funguilds$guild[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_conf <- funguilds$confidenceRanking[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_growthform <- funguilds$growthForm[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_trait <- funguilds$trait[match(fbp$my_otu, funguilds$OTU)]
fbp$fung_notes <- funguilds$notes[match(fbp$my_otu, funguilds$OTU)]

fb_pos_assigned <- fbp[!is.na(fbp$taxlevel),]

#table(fb_pos_assigned$fung_trophicMode)/nrow(fb_pos_assigned)
table(fb_pos_assigned$fung_trophicMode)/length(fb_fung_pos)

# what proportion of fungal nodes could be assigned guild
dim(fb_pos_assigned)[1] / dim(fbp)[1]



## fungal-fungal edges
fung_only <- x[grepl("fung", x$from) & grepl("fung", x$to),]

fonly_neg <- fung_only[fung_only$sign=="negative",]

fonly_fung_neg <- unique(c(fonly_neg$from, fonly_neg$to))

fneg <- data.frame(my_otu = fonly_fung_neg)

fneg$taxonomy <- funguilds$Taxonomy[match(fneg$my_otu, funguilds$OTU)]
fneg$taxlevel <- funguilds$taxon[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_trophicMode <- funguilds$trophicMode[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_guild <- funguilds$guild[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_conf <- funguilds$confidenceRanking[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_growthform <- funguilds$growthForm[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_trait <- funguilds$trait[match(fneg$my_otu, funguilds$OTU)]
fneg$fung_notes <- funguilds$notes[match(fneg$my_otu, funguilds$OTU)]

fonly_neg_assigned <- fneg[!is.na(fneg$taxlevel),]

table(fonly_neg_assigned$fung_trophicMode)/nrow(fonly_neg_assigned)
table(fonly_neg_assigned$fung_trophicMode)/nrow(fonly_neg)

# what proportion of fungal nodes could be assigned guild
dim(fonly_neg_assigned)[1] / dim(fneg)[1]



fonly_pos <- fung_only[fung_only$sign=="positive",]

fonly_fung_pos <- unique(c(fonly_pos$from, fonly_pos$to))

fpos <- data.frame(my_otu = fonly_fung_pos)

fpos$taxonomy <- funguilds$Taxonomy[match(fpos$my_otu, funguilds$OTU)]
fpos$taxlevel <- funguilds$taxon[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_trophicMode <- funguilds$trophicMode[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_guild <- funguilds$guild[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_conf <- funguilds$confidenceRanking[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_growthform <- funguilds$growthForm[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_trait <- funguilds$trait[match(fpos$my_otu, funguilds$OTU)]
fpos$fung_notes <- funguilds$notes[match(fpos$my_otu, funguilds$OTU)]

fonly_pos_assigned <- fpos[!is.na(fpos$taxlevel),]

table(fonly_pos_assigned$fung_trophicMode)/nrow(fonly_pos_assigned)
table(fonly_pos_assigned$fung_trophicMode)/nrow(fonly_pos)

# what proportion of fungal nodes could be assigned guild
dim(fonly_pos_assigned)[1] / dim(fbp)[1]

```

```{r}
fung_tax <- readRDS("../intermediates/global/fung_seed_taxonomy.rds")

fung_new_tax <- data.frame("OTU ID" = fung_tax$denovo_OTU)
names(fung_new_tax) <- "OTU ID"

fung_new_tax$Kingdom <- paste0("k__", fung_tax$Kingdom)
fung_new_tax$Phylum <- paste0("p__", fung_tax$Phylum)
fung_new_tax$Class <- paste0("c__", fung_tax$Class)
fung_new_tax$Order <- paste0("o__", fung_tax$Order)
fung_new_tax$Family <- paste0("f__", fung_tax$Family)
fung_new_tax$Genus <- paste0("g__", fung_tax$Genus)
fung_new_tax$Species <- paste0("s__", fung_tax$Species)


fung_new_tax$paste_tax <- paste(fung_new_tax$Kingdom, fung_new_tax$Phylum, fung_new_tax$Class, fung_new_tax$Order, fung_new_tax$Family, fung_new_tax$Genus, fung_new_tax$Species, sep=";")


bact_tax <- readRDS("../intermediates/global/bact_seed_taxonomy.rds")

bact_new_tax <- data.frame("OTU ID" = bact_tax$denovo_OTU)
names(bact_new_tax) <- "OTU ID"

bact_new_tax$Kingdom <- paste0("k__", bact_tax$Kingdom)
bact_new_tax$Phylum <- paste0("p__", bact_tax$Phylum)
bact_new_tax$Class <- paste0("c__", bact_tax$Class)
bact_new_tax$Order <- paste0("o__", bact_tax$Order)
bact_new_tax$Family <- paste0("f__", bact_tax$Family)
bact_new_tax$Genus <- paste0("g__", bact_tax$Genus)
bact_new_tax$Species <- paste0("s__", bact_tax$Species)


bact_new_tax$paste_tax <- paste(bact_new_tax$Kingdom, bact_new_tax$Phylum, bact_new_tax$Class, bact_new_tax$Order, bact_new_tax$Family, bact_new_tax$Genus, bact_new_tax$Species, sep=";")
```

```{r}
fung_genus_assigned <- fung_new_tax[!grepl("NA", fung_new_tax$Genus),]

bact_genus_assigned <- bact_new_tax[!grepl("ceae", bact_new_tax$Genus),]
bact_genus_assigned <- bact_genus_assigned[!grepl("unclassified", bact_genus_assigned$Genus),]
bact_genus_assigned <- bact_genus_assigned[!grepl("uncultured", bact_genus_assigned$Genus),]

cand_alb$tax <- ifelse(grepl("fung", cand_alb$to), 
                       fung_new_tax$paste_tax[match(cand_alb$to, fung_new_tax$`OTU ID`)],
                       bact_new_tax$paste_tax[match(cand_alb$to, bact_new_tax$`OTU ID`)])

allnodes <- unique(c(x$from, x$to))

all_fung <- allnodes[grepl("fung", allnodes)]
all_bact <- allnodes[grepl("bact", allnodes)]

fung_keep <- all_fung[all_fung %in% fung_genus_assigned$`OTU ID`]

bact_bact <- x[(grepl("bact", x$from) & grepl("bact", x$to)),]

```



```{r}
no_fam_fonlyneg <- fonly_neg_assigned[!(grepl("ceae", fonly_neg_assigned$taxonomy)),]

no_fam_fbneg <- fb_neg_assigned[!(grepl("ceae", fb_neg_assigned$taxonomy)),]

# candida albicans
candida_alb <- x[x$from=="fung_otu23168",]
candida_alb2 <- x[x$to=="fung_otu23168",]
candida_alb2$to <- candida_alb2$from
candida_alb2$from <- "fung_otu23168"

cand_alb <- rbind(candida_alb, candida_alb2)

cand_alb$tax <- ifelse(grepl("fung", cand_alb$to), fung_new_tax$paste_tax[match(cand_alb$to, fung_new_tax$`OTU ID`)], bact_new_tax$paste_tax[match(cand_alb$to, bact_new_tax$`OTU ID`)])

# fusidium
fus <- x[x$from=="fung_otu5766",]
fus2 <- x[x$to=="fung_otu5766",]
fus2$to <- fus2$from
fus2$from <- "fung_otu5766"

fusidium <- rbind(fus, fus2)
fusidium$tax <- ifelse(grepl("fung", fusidium$to), fung_new_tax$paste_tax[match(fusidium$to, fung_new_tax$`OTU ID`)], bact_new_tax$paste_tax[match(fusidium$to, bact_new_tax$`OTU ID`)])


# Phialemoniopsis pluriloculosa
plu <- x[x$from %in% c("fung_otu8550", "fung_otu26250", "fung_otu13241"),]
plu2 <- x[x$to %in% c("fung_otu8550", "fung_otu26250", "fung_otu13241"),]
plu2$to <- plu2$from

plurilo <- rbind(plu, plu2)
plurilo$tax <- ifelse(grepl("fung", plurilo$to), fung_new_tax$paste_tax[match(plurilo$to, fung_new_tax$`OTU ID`)], bact_new_tax$paste_tax[match(plurilo$to, bact_new_tax$`OTU ID`)])  

# acremonium 
aa <- no_fam_fonlyneg$my_otu[grepl("Acremonium", no_fam_fonlyneg$taxlevel)]
acr <- x[x$from %in% aa,]
acr2 <- x[x$to %in% aa,]
acr2$from <- "fung_otu18257"

acremon <- rbind(acr, acr2)
acremon$tax <- ifelse(grepl("fung", acremon$to), 
                      fung_new_tax$paste_tax[match(acremon$to, fung_new_tax$`OTU ID`)],
                      bact_new_tax$paste_tax[match(acremon$to, bact_new_tax$`OTU ID`)])  

acremon$tax[grepl("Asper", acremon$tax)]

# tolypocladium 
tp <- funguilds$OTU[grepl("Tolypocladium", funguilds$Taxonomy)]
tol <- x[x$from %in% tp,]
tol2 <- x[x$to %in% tp,]
tol2$from <- "fung_otu12042"

tolypo <- rbind(tol, tol2)
tolypo$tax <- ifelse(grepl("fung", tolypo$to), 
                      fung_new_tax$paste_tax[match(tolypo$to, fung_new_tax$`OTU ID`)],
                      bact_new_tax$paste_tax[match(tolypo$to, bact_new_tax$`OTU ID`)])  

tolypo$tax[grepl("Elaph", tolypo$tax)]


# trichoderma 
td <- funguilds$OTU[grepl("Trichoderma", funguilds$Taxonomy)]
tri <- x[x$from %in% td,]
tri2 <- x[x$to %in% td,]
tri2$from <- "fung_otu12042"

tricho <- rbind(tri, tri2)
tricho$tax <- ifelse(grepl("fung", tricho$to), 
                      fung_new_tax$paste_tax[match(tricho$to, fung_new_tax$`OTU ID`)],
                      bact_new_tax$paste_tax[match(tricho$to, bact_new_tax$`OTU ID`)])  

tricho$tax[grepl("Elaph", tricho$tax)]

# weissella 
ws <- bact_new_tax$`OTU ID`[grepl("Weissella", bact_new_tax$paste_tax)]
wei <- x[x$from %in% ws,]
wei2 <- x[x$to %in% ws,]
wei2$from <- "bact_otu43197"

weiss <- rbind(wei, wei2)
weiss$tax <- ifelse(grepl("fung", weiss$to), 
                      fung_new_tax$paste_tax[match(weiss$to, fung_new_tax$`OTU ID`)],
                      bact_new_tax$paste_tax[match(weiss$to, bact_new_tax$`OTU ID`)])  

weiss$tax[grepl("Elaph", weiss$tax)]

# streptococcus 
y = igraph::as_data_frame(filt_bact_net)

sc <- bact_new_tax$`OTU ID`[grepl("Streptococcus", bact_new_tax$paste_tax)]
str <- y[y$from %in% sc,]
str2 <- y[y$to %in% sc,]
str2$from <- "bact_otu12182"

strep <- rbind(str, str2)
strep$tax <- ifelse(grepl("fung", strep$to), 
                      fung_new_tax$paste_tax[match(strep$to, fung_new_tax$`OTU ID`)],
                      bact_new_tax$paste_tax[match(strep$to, bact_new_tax$`OTU ID`)])  

strep$tax[grepl("Elaph", strep$tax)]
```


### Sample data table: Fungal guild assignments for fungal nodes participating in negative fungal-bacterial edges

```{r, echo=FALSE}
rmarkdown::paged_table(fb_neg_assigned[-2])
```

### What proportion of all fungal nodes participating in negative fungal-bacterial edges could be assigned guild?
All combinations tested (negative fungal-bacterial edges, positive fungal-bacterial, negative fungal-fungal, positive fungal-fungal) resulted in around this level of guild assignment
```{r, echo=FALSE}
# what proportion of fungal nodes could be assigned guild
dim(fb_neg_assigned)[1] / dim(fbn)[1]
```

### Proportions of guilds in negative vs positive edges (fungal-bacterial edges)
 
Fungi in **negative** fungal-bacterial edges
```{r, echo=FALSE}
table(fb_neg_assigned$fung_trophicMode)/length(fb_fung_neg)
```

Fungi in **positive** fungal-bacterial edges
```{r, echo=FALSE}
table(fb_pos_assigned$fung_trophicMode)/length(fb_fung_pos)
```

### Proportions of guilds in negative vs positive edges (fungal-fungal edges)
Some differences exist but we're always only accounting for <50% of fungal OTUs
<br>

Fungi in **negative** fungal-fungal edges
```{r, echo=FALSE}
table(fonly_neg_assigned$fung_trophicMode)/nrow(fonly_neg)
```

Fungi in **positive** fungal-fungal edges
```{r, echo=FALSE}
table(fonly_pos_assigned$fung_trophicMode)/nrow(fonly_pos)
```


### A lot of overlap in the fungi that participate in negative vs positive edges
```{r}
length(fonly_fung_neg) # fungi in negative f-f edges
length(fonly_fung_pos) # fungi in positive f-f edges

sum(fonly_fung_neg %in% fonly_fung_pos) / length(fonly_fung_neg) # all fungi in neg edges are present in pos edge fungi


length(fb_fung_neg) # fungi in negative f-b edges
length(fb_fung_pos) # fungi in positive f-b edges

sum(fb_fung_neg %in% fb_fung_pos) / length(fb_fung_neg) # most fungi in neg edges are present in pos edge fungi

```

### 23 OTUs only take part in negative fung-bact edges (are not present in positive f-b edges). These are their guild assignments (for those that could be assigned).

```{r, echo=FALSE}
neg_check <- setdiff(fb_fung_neg, fb_fung_pos)

chk_df <- fb_neg_assigned[fb_neg_assigned$my_otu %in% neg_check,]

rmarkdown::paged_table(chk_df[,-2])

```




